Can historical data predict population responses to climate change experiments?

### 2016-12-22 precip/manuscript/precip_ms.tex
edits to the results

The continuously recorded soil moisture data also showed treatment effects on soil moisture, but these effects were weaker on average than the spot measurements and depended on season and recent rainfall (table \ref{table:soil_moisture_model}; fig \ref{fig:dailyVWC}). We saw weaker effects during the spring than during the fall and summer: the drought plots were about 20-30\% drier in the fall and summer but only 7 to 14\% drier during the spring, while the irrigated plots were 30\% wetter during the fall and summer but only 20-25\% wetter during the spring.  Treatment differences were slightly larger during rainy periods, especially in the spring. We did not find evidence that the drought shelters and the irrigation treatments consistently affected air temperature at 30 cm above the plots.

The SOILWAT soil moisture model predicted average monthly soil volumetric water content of between 10 ml/ml and 15 ml/ml each month, with the month of April being the wettest and the month of July, August and September being the driest on average. Annual variation in seasonal soil moisture for each year was positively correlated with seasonal precipitation and negatively correlated with seasonal temperature. During the course of the experiment, SOILWAT reproduced much of the daily variation observed in soil moisture recorded by our automatic data loggers, but the average soil moisture predicted by SOILWAT was about 5 ml/ml wetter than the soil moisture content recorded by the data loggers.

After adjusting the SOILWAT seasonal soil moisture predictions by the treatment effects, we found that the soil moisture predicted in the drought plots during the course of the experiment was well below the historical seasonal averages: the summer of 2012 and 2013, the fall of 2013, and the spring and winter of 2014 fell below the 5th percentile limit for drought in the historical period \ref{fig:seasonalVWC}. Soil moisture in our irrigation plots was generally above the historical average soil moisture but conditions never exceeded the 90th percentile for soil moisture in the historical period.

@@

The cover of \textit{H. comata} and \textit{P. spicata} fell significantly in the drought plots from 2011 to 2016 (tables \ref{table:changeHECO}, \ref{table:changePSSP}, fig \ref{fig:coverChange}). The cover of \textit{P. secunda} showed a slight decrease in the drought plots and an increase in the irrigated plots but these changes were not significant (table \ref{table:changePOSE}).  In contrast to the grasses, the cover of \textit{A. tripartita} increased slightly in all three treatments (fig \ref{fig:coverChange}).

Our treatment models fit to the experimental and observational data indicated a variety of treatment effects on the vital rates of each species. After comparing WAIC scores for the growth models with and without the size by treatment effects, we retained size by treatment interaction effects in the growth models for \textit{A. tripartita} and \textit{P. secunda}, and the survival model for \textit{P. secunda}. For \textit{A. tripartita} we found significant size by treatment effects of drought: drought had positive effects on plants of average size and smaller \ref{fig:growthTreat}, but plants larger than the mean size by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls. \textit{A. tripartita} showed the opposite response in the irrigated plots, (although the irrigation parameters were not technically significant at the 95\% confidence level): irrigation reduced growth for small plants while irrigation increased growth of plants more than 1.5 standard deviations greater than the mean size. Drought led to a strong (but not signficant) decrease in \textit{H. comata} growth, while irrigation had no effect on growth.  Like \textit{A. tripartita}, we saw size by treatment effects on \textit{P.secunda} growth, with the negative effects of drought becoming greater for larger plants. \textit{P. secunda} showed the opposite response in the irrigation plots with larger plants showing the largest increase in growth in response to irrigation (although these effects were technically not significant). \textit{P. spicata} growth was relatively unaffected by the drought and irrigation treatments.

Drought decreased the survival of all three grass species (fig \ref{fig:survivalTreat}). And \textit{P. secunda} showed a negative size by drought interaction effect, indicating that the survival of larger plants was more negatively affected by drought than that of the smaller plants, with the smallest plants (plants one standard deviation smaller than the mean) actually seeing a slightly positive effect of drought. \textit{A. tripartita} survival was relatively unaffected by the drought and irrigation treatments.

Recruitment in our irrigation plots was significantly less than in control plots for two grass species \textit{P. secunda} and \textit{P. spicata} (fig \ref{fig:recruitmentTreat}). However, recruitment was also lower in the drought plots than in the the control plots (although not significantly so), indicating that the decrease in the irrigated plots may have not been entirely due to the irrigation itself. The recruitment data for \textit{A. tripartita} were relatively limited, with only 32 new recruits in total observed in all 30 plots over the course of the five year experiment.

As expected from previous research most of our demographic models estimated strong negative intra-specific crowding effects and weaker negative inter-specific crowding effects on the focal species (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015}.

@@

We choose three seasonal soil moisture variables for each species' climate model based on their correlation with the random year effects in the no climate model fit to the observational data (table \ref{table:strongCor}). We included size by soil moisture variables for \textit{A. tripartita} and  \textit{P. secunda} based on the treatment response we observed in the experiment. All three time lags and all three seasons show up in the choosen variables. After fitting the vital rate models with the selected soil moisture variable we observed a trend towards positive soil moisture effects on growth of all three grasses \ref{fig:climateGrowth}. For \textit{H.comata} the soil moisture of the most recent summer had a significantly positive effect while the soil moisture of the previous summer and the fall before that were also positive but not significant. For \textit{A. tripartita} the summer and fall soil moisture of the previous year had strong negative effects on growth.  There were also strong positive size by climate interaction effects for these variables: soil moisture had a stronger negative effect on small plants and a positive effect only on the largest plants.

Soil moisture had significant effects on the survival of all four species \ref{fig:climateSurvival}. As for growth the grasses showed mainly positive effects while \textit{A. tripartita} showed a significant negative effect of soil moisture in the summer of the previous year and a strong negative effect of spring soil moisture of the previous year. \textit{H. comata} showed a significant positive effect of spring lag soil moisture and a strong positive effect of spring soil moisture in the previous year. \textit{P. secunda} showed a significant positive effect of the previous spring's soil moisture and there was a significant interaction between this effect and plant size: as plant size increased this effect became more strongly positive. Finally for \textit{P. spicata} there was a significant positive effect of lag spring soil moisture on survival.

There were only two significant effects of soil moisture on recruitment: lag fall soil moisture had a positive effect on \textit{P. secunda}, and lag summer soil moisture had a negative effect on \textit{P. spicata} recruitment. Soil moisture of the previous year's summer had also had a strong negative effect on \textit{P. spicata} recruitment.

The intra- and interspecific crowding effects estimated in the climate model were similar to those estimated in the treatment model (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015}.

@@

For most models adding climate covariates did not improve our ability to predict species responses in the experiment \ref{table:overallPreds}. However, the climate models did improve overall prediction MSE for growth of \textit{A. tripartita} and growth and survival of \textit{P. secunda} (table \ref{table:overallPreds}). In terms of lppd, the climate model outperformed the no climate model for \textit{A. tripartita} growth, \textit{H. comata} recruitment, \textit{P. secunda} growth and survival and \textit{P. spicata} recruitment.  .

When we look at the predictions for each treatment separately we see that climate covariates improved model predictions more often in the drought treatments than in the control or irrigation treatments \ref{table:treatmentPreds}. For all four species, the climate model outperformed the no climate model for predicting the response of growth to drought in terms of lppd \ref{table:treatmentPreds}. The climate model outperformed the no climate model for predicting irrigation effects on growth for all species except \textit{H. comata}.

Overall our climate models often predicted the correct direction of the drought and irrigation treatments \ref{fig:parPredictions}. In four cases we both observed and predicted treatment effects significantly different from zero based on the 95\% Bayesian credible interval around the parameters: the drought response of \textit{H. comata} survival (fig \ref{fig:parPredHECOSurvival}), the drought response of \textit{P. secunda} growth (fig \ref{fig:parPredPOSEGrowth}), the irrigation response of \textit{P spicata} recruitment (fig \ref{fig:parPredPSSPRecruitment}) and the irrigation response of \textit{P. secunda} recruitment (fig \ref{fig:parPredPOSERecruitment}).  In only one of these cases, for \textit{P. secunda} recruitment, was the predicted effect in the opposite direction from the observed treatment effect \ref{fig:parPredictions}. The overall correlation between the predicted and observed treatment effects for all treatments, species and vital rates was r = 0.54, whereas the correlation for the drought treatment effects (r = 0.77) was better than for the irrigation effects (r = 0.46).  Also for the three models in which we included size by treatment or size by climate interactions the correlation of these size dependent effects were much stronger than the intercept parameter estimates \ref{fig:parPredictions}.

Using the vital rate models for each species we generated one step ahead cover predictions for each quadrat in each year of the experiment.  Average cover predicted by the climate model tended to be lower than the observed cover each year for \textit{A. tripartita} and \textit{P.secunda} (fig \ref{fig:coverPred}). Comparing the overall population growth rates predicted to those observed in the experiment, we see that the MSE of the climate model was lower than the MSE of the no climate model for \textit{P. secunda} and \textit{P. spicata} (table \ref{table:corPGR}). The predictions produced by the climate model for these species were also slightly more correlated with the observations than the predictions produced by the no climate model (table \ref{table:corPGR}). Considering each treatment and species separately we see that the predicted population growth rates for \textit{A. tripartita}, \textit{P. secunda} and \textit{P. spicata} were all consistently lower than the observed population growth rates (figs \ref{fig:pgrARTR}, \ref{fig:pgrPOSE}, \ref{fig:pgrPSSP}). The climate model showed lower MSE for \textit{A. tripartita}, \textit{P. secunda} and \textit{P. spicata} in the irrigation treatment, \textit{P. spicata} in the control treatment and \textit{H. comata} in the drought treatment (figs \ref{fig:pgrHECO}).  However, the correlations between the predicted and observed log changes in cover did not always show the same pattern as MSE: the climate model made more strongly correlated predictions with the observations than the no climate model only for \textit{P. spicata} and \textit{P. secunda} in the control treatment and \textit{P. secunda} and \textit{H. comata} in the drought treatment.

@@

Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species response to experimental climate manipulations \ref{fig:parPredictions}. This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models \ref{table:overallPreds}. This should give us some hope that even when climate effects in demographic models to observational data are weak, they may contain useful qualitative information on the direction of climate effects in the future.

@@

Funding was provided by an NSF GRFP to AK, NSF grants DEB-1353078, and DEB-1054040 to PBA and by the Utah Agricultural Experiment Station (get journal paper number). The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site. Joe and Kevin, Lau, John Bradford, Caitlin (soilwat),

>>>>

The continuously recorded soil moisture data also showed treatment effects, but these were weaker on average than the spot measurements and depended on season and recent rainfall (table \ref{table:soil_moisture_model}; fig \ref{fig:dailyVWC}). We saw weaker effects during the spring than during the fall and summer: the drought plots were about 20-30\% drier than ambient in the fall and summer but only 7 to 14\% drier during the spring, while the irrigated plots were 30\% wetter during the fall and summer but only 20-25\% wetter during the spring.  Treatment differences were slightly larger during rainy periods, especially in the spring (rainfall effect in table \ref{table:soil_moisture_model}). We did not find evidence that the drought shelters and the irrigation treatments consistently affected air temperature at 30 cm above the plots.

The SOILWAT soil moisture model predicted average monthly soil volumetric water content of between 10 ml/ml and 15 ml/ml each month, with the month of April being the wettest and the month of July, August and September being the driest on average. Annual variation in seasonal soil moisture for each year was positively correlated with seasonal precipitation and negatively correlated with seasonal temperature. During the course of the experiment, SOILWAT reproduced much of the daily variation observed in soil moisture recorded by our automatic data loggers, but the average soil moisture predicted by SOILWAT was about 5 ml/ml higher than the soil moisture content observed in the field.

After adjusting the SOILWAT seasonal soil moisture predictions by the treatment effects, we found that the soil moisture predicted in the drought plots during the course of the experiment was well below the historical seasonal averages: the summer of 2012 and 2013, the fall of 2013, and the spring and winter of 2014 fell below the 5th percentile limit for drought in the historical period (fig \ref{fig:seasonalVWC}). Soil moisture in our irrigation plots was generally above the historical average soil moisture but conditions never exceeded the 90th percentile for soil moisture in the historical period (fig \ref{fig:seasonalVWC}).

@@

The cover of \textit{H. comata} and \textit{P. spicata} fell significantly in the drought plots from 2011 to 2016 (tables \ref{table:changeHECO}, \ref{table:changePSSP}; fig \ref{fig:coverChange}). The cover of \textit{P. secunda} showed a slight decrease in the drought plots and an increase in the irrigated plots but these changes were not significant (table \ref{table:changePOSE}).  In contrast to the grasses, the cover of \textit{A. tripartita} increased slightly in all three treatments (fig \ref{fig:coverChange}).

Our treatment models fit to the experimental and observational data indicated a variety of treatment effects on the vital rates of each species. Based on the WAIC scores with and without the size by treatment effects, we retained size by treatment effects in the growth models for \textit{A. tripartita} and \textit{P. secunda}, and the survival model for \textit{P. secunda}. For \textit{A. tripartita} we found significant size by treatment effects of drought: drought had positive effects on plants of average size and smaller (fig \ref{fig:growthTreat}), but plants larger than the mean size by more than 1.5 standard deviations grew slightly less in the drought treatment than in the controls. \textit{A. tripartita} showed the opposite response in the irrigated plots, (although the irrigation parameters were not significant at the 95\% confidence level): irrigation reduced growth for small plants while irrigation increased growth of plants more than 1.5 standard deviations larger than the mean size. Drought led to a strong (but not signficant) decrease in \textit{H. comata} growth, while irrigation had no effect on growth.  Like \textit{A. tripartita}, we saw size by treatment effects on \textit{P.secunda} growth, with the negative effects of drought becoming greater for larger plants. \textit{P. secunda} showed the opposite response in the irrigation plots with larger plants showing the largest increase in growth in response to irrigation (although not significant). \textit{P. spicata} growth was relatively unaffected by the drought and irrigation treatments.

Survival of all three grass species (fig \ref{fig:survivalTreat}) decreased in the drought plots. And \textit{P. secunda} showed a negative size by drought interaction effect: the survival of larger plants was more negatively affected by drought than that of the smaller plants. \textit{A. tripartita} survival was relatively unaffected by the drought and irrigation treatments.

Recruitment in our irrigation plots was significantly less than in control plots for two grass species \textit{P. secunda} and \textit{P. spicata} (fig \ref{fig:recruitmentTreat}). However, recruitment was also lower in the drought plots than in the the control plots (although not significantly so), indicating that the decrease in the irrigated plots may have not been entirely due to the irrigation itself. The recruitment data for \textit{A. tripartita} were relatively limited, with only 32 new recruits in total observed in all 30 plots over the course of the five year experiment and we observed no treatment effects.

Consistent with previous research most of our demographic models estimated strong negative intra-specific crowding effects and weaker negative inter-specific crowding effects on the focal species (appendix) \citep{adler_coexistence_2010,chu_direct_2016,chu_large_2015,adler_weak_2016}.

@@

We choose three seasonal soil moisture variables for each species' based on their correlation with the random year effects in the no climate model (table \ref{table:strongCor}). We included size by soil moisture variables for \textit{A. tripartita} and  \textit{P. secunda} based on the treatment response we observed in the experiment. All three time lags and all three seasons show up in the selected variables. After fitting the vital rate models with the selected soil moisture variable we observed a trend towards positive soil moisture effects on growth of all three grasses (fig \ref{fig:climateGrowth}). For \textit{H.comata} the soil moisture of the most recent summer ($summer_1$) had a significantly positive effect while the soil moisture during $summer_0$ and $fall_lag$ were also positive but not significant. For \textit{A. tripartita} $fall_0$ and $summer_0$ had strong negative effects on growth. There were also strong positive size by climate interaction effects for these variables: soil moisture had a stronger negative effect on small plants and a positive effect only on the largest plants (fig \ref{fig:parPredARTRGrowth}).

Soil moisture had significant effects on the survival of all four species (fig  \ref{fig:climateSurvival}). As for growth the grasses showed mainly positive effects while \textit{A. tripartita} showed a significant negative effect of $summer_0$ and a strong negative effect of $spring_0$. \textit{H. comata} showed a significant positive effect of $spring_lag$ soil moisture and a strong positive effect of $spring_0$ and $fall_1$. \textit{P. secunda} showed a significant positive effect of the previous $spring_0$ and there was an interaction between this effect and plant size: as plant size increased this effect became more positive. Finally for \textit{P. spicata} there was a significant positive effect of $spring_lag$ soil moisture on survival.

There were only two significant effects of soil moisture on recruitment: $fall_lag$ soil moisture had a positive effect on \textit{P. secunda}, and $summer_lag$ soil moisture had a negative effect on \textit{P. spicata} recruitment (fig \ref{fig:climateRecruitment}). Soil moisture of $summer_0$ also had a strong negative effect on \textit{P. spicata} recruitment.

The intra- and interspecific crowding effects estimated in the climate model were similar to those estimated in the treatment model (appendix).

@@

For most models adding climate covariates did not improve our ability to predict species responses in the experiment (table \ref{table:overallPreds}). However, the climate models did improve overall prediction MSE for growth of \textit{A. tripartita} and growth and survival of \textit{P. secunda} (table \ref{table:overallPreds}). In terms of lppd, the climate model outperformed the no climate model in six out of twelve models: for \textit{A. tripartita} growth, \textit{H. comata} recruitment, \textit{P. secunda} growth and survival and \textit{P. spicata} recruitment (table \ref{table:overallPreds}).

When we look at the predictions for each treatment separately we see that climate covariates improved model predictions more often in the drought treatments than in the control or irrigation treatments (table \ref{table:treatmentPreds}). For all four species, the climate model outperformed the no climate model for predicting the response of growth to drought in terms of lppd (table \ref{table:treatmentPreds}). The climate model also outperformed the no climate model for predicting irrigation effects on growth for all species except \textit{H. comata}.

Overall our climate models often predicted the correct direction of the drought and irrigation treatments (fig \ref{fig:parPredictions}). In four cases we both observed and predicted treatment effects significantly different from zero based on the 95\% Bayesian credible interval around the parameter mean: the drought response of \textit{H. comata} survival (fig \ref{fig:parPredHECOSurvival}), the drought response of \textit{P. secunda} growth (fig \ref{fig:parPredPOSEGrowth}), the irrigation response of \textit{P spicata} recruitment (fig \ref{fig:parPredPSSPRecruitment}) and the irrigation response of \textit{P. secunda} recruitment (fig \ref{fig:parPredPOSERecruitment}).  In only one of these cases, for \textit{P. secunda} recruitment, was the predicted effect in the opposite direction from the observed treatment effect (fig \ref{fig:parPredictions}). The overall correlation between the predicted and observed treatment effects for all treatments, species and vital rates was r = 0.54, whereas the correlation for the drought treatment effects, r = 0.77, was better than for the irrigation effects, r = 0.46.  Also the correlation between the size by climate and size by treatment effects for \textit{A. tripartita} growth and \textit{P. secunda} growth and survival was much stronger than the correlation between the intercept parameter estimates (fig \ref{fig:parPredictions}).

Using the vital rate models for each species we generated one step ahead cover predictions for each quadrat in each year of the experiment.  Average cover predicted by the climate model tended to be lower than the observed cover each year for \textit{A. tripartita} and \textit{P.secunda} (fig \ref{fig:coverPred}). Comparing the overall population growth rates predicted to those observed in the experiment, we see that the MSE of the climate model was lower than the MSE of the no climate model for \textit{P. secunda} and \textit{P. spicata} (table \ref{table:corPGR}). The predictions produced by the climate model for these species were also slightly more correlated with the observations than the predictions produced by the no climate model (table \ref{table:corPGR}). Considering each treatment and species separately, the predicted population growth rates for \textit{A. tripartita}, \textit{P. secunda} and \textit{P. spicata} were all consistently lower than the observed population growth rates (figs \ref{fig:pgrARTR}, \ref{fig:pgrPOSE}, \ref{fig:pgrPSSP}). The climate model showed lower MSE for \textit{A. tripartita}, \textit{P. secunda} and \textit{P. spicata} in the irrigation treatment, \textit{P. spicata} in the control treatment and \textit{H. comata} in the drought treatment (fig \ref{fig:pgrHECO}).  However, the correlations between the predicted and observed log changes in cover did not always show the same pattern as MSE: the climate model made more strongly correlated predictions with the observations than the no climate model only for \textit{P. spicata} and \textit{P. secunda} in the control treatment and \textit{P. secunda} and \textit{H. comata} in the drought treatment.

@@

Our experiment showed that observational data on the response of plant populations to interannual climate variation can indeed help us predict the direction of species responses to experimental climate manipulations (fig \ref{fig:parPredictions}). This was true even though adding climate parameters to the demographic models only improved vital rate predictions for half of the models (fig \ref{table:overallPreds}). This should give us some hope that even when climate effects in demographic models fit to observational data are weak, they may contain useful qualitative information on the direction of climate effects in the future.

@@

Funding was provided by NSF grants DEB-1353078, and DEB-1054040 to PBA, an NSF GRFP award to AK, and by the Utah Agricultural Experiment Station (get journal paper number). The USDA-ARS Sheep Experiment Station generously provided access to historical data and the field experiment site. Laureano Gherardi provided us with valuable advice in constructing the rain-out and irrigation experiment.
